---
title: "Basic Statistics: Mean, Median, Standard Deviation, Variance, and Samples"
author: "Jack Hester"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Mean

#### Definition

The mean, or average, of a data set (set of observations) is calculated as:

$$\overline{x}=\frac{\sum_{i=1}^{n} x_{i}}{n}$$
where $\overline{x}$ is the (sample) mean, $x_{i}$ is individual observation $i$, $n$ is the total number of samples (observations), and $\sum_{i=1}^{n} x_{i}$ is the sum of all observations, which can be expanded as $\sum_{i=1}^{n} x_{i}=x_{1}+x_{2}+x_{3}+\ldots \ldots x_{n}$.

####  Calculating mean in R
```{r}
x <- c(0.6,11,7,6,3.2,-100) #your set of values (observations)
mean(x)
```

#### Why is it sometimes called the SAMPLE mean?

When we don't have a value for every person/item/thing in a population (group), we use a $\it{sample}$ that hopefully represents the whole population. When that's the case, we say it's the sample mean to distinguist between it from the population mean, wich is not known with certainty since we cannot measure all of its values.

Typically, $\overline{x}$ represents the sample mean, and $\mu$ represents the (usually unknown) population mean

## Median

The median is the middle value of an \em{ordered} (in order from least to greatest) set of continuous data measures/observations. For example, the median of the numbers 75, 90, 120, 121, 140 is 120. If there is an even number of observations, such as 20, 33, 36, 41 then you take the average (mean) of the two values in the middle; here that's (33+36)/2 = 34.5. 

#### Calculating mean in R
```{r}
x <- c(0.6,11,7,6,3.2,-100) #your set of values (observations)
median(x)
```

#### Outliers: median versus mean
The median is more resistent to outliers (extreme values on one side or the other of the data set). Why? Let's think about an example...

Let's start with the set of observations (1.5,1.55,1.7,1.71,1.9,1.92,1.98).
The sample mean $\overline{x}$ is 1.75, and the mean is 1.71.

Now let's add an outlier to the data, and make the set of observations (1.5,1.55,1.7,1.71,1.9,1.92,1.98,5).

The new sample mean is 2.16 (a large jump from 1.75), but the median is 1.81, much closer to the original mean and median values...

In general, it's reasonably likely, especially as the number of samples increases, that values towards the middle of the set of samples/observations (again, assuming it's ordered) will be fairly close together and not contain outliers. As illustrated in the example, this means that the median will not shift that far away from the "middle" with the addition of an outlier, whereas the mean might shift much more (especially with smaller sample sizes).

##Variance

Variance is the average of the square of the deviations about the sample mean. The equation for (sample) variance is:

$$s^{2}=\frac{\sum_{i=1}^{n}\left(x_{i}-\overline{x}\right)^{2}}{n-1}$$

## Standard Deviation
